Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-25 article/article.tex
Commit 2024-06-25 21:11

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

We studied the enhancement of the global horizontal irradiance induced
by clouds over Thessaloniki for the period 1993 -- 2023 using data
recorded every one minute. We identified the cloud enhancement events
(CE) by creating an appropriate clear sky irradiance reference with the
use of a radiative transfer model and aerosol optical depth data from a
collocated Cimel sun photometer and a Brewer spectrophotometer. We found
a trend in CE events of \(45.6\pm 21.9\,\text{cases}/\text{year}\), and
a trend in the CE events irradiation of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The peak of the CE events was
observed during May and June. An analysis of the total duration of CE
events, showed that durations longer than 5 minutes are very rare, with
exceptions lasting over an hour. Finally we have detected enhancements
above the total solar irradiance at the top of the atmosphere of up to
\(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
\(200\,\text{W}/\text{m}^{2}\).

CE: cloud enhancement cases one minute

ECE: extreme cloud enhancement cases one minute over TSI on horizontal
plane

CEG: cloud enhancement groups, cases events with consecutive CE

\(\text{GHI}_\text{i}\): measured one minute global horizontal
irradiance

\(\text{GHI}_\text{CSm}\): modelled clear sky one minute global
horizontal irradiance

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

In this study, we evaluate the effects CE on GHI by investigating their
frequency of occurrence, intensity, and duration in a thirty-year period
of GHI observations at Thessaloniki, Greece. We used modeled clear sky
irradiance as a baseline to identify cloud enhancements and we
determined long-term trends of the above mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

\hypertarget{data-and-methodology}{%
\section{Data and methodology}\label{data-and-methodology}}

\hypertarget{instrumentation-and-data}{%
\subsection{Instrumentation and data}\label{instrumentation-and-data}}

@@

\hypertarget{cloud-enhancement-detection}{%
\subsection{Cloud enhancement
detection}\label{cloud-enhancement-detection}}

@@

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}, and the total solar irradiance (TSI) at
the top of the atmosphere, adjusted for the Sun-Earth distance. We have
tested both methods by using an appropriate relative threshold and/or an
additional constant offset. The initial results showed that we can
detect a big portion of the CE events. However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric
conditions introduced numerous false positive and false negative
results. The main reason for these discrepancies is the variability of
the effects of aerosols and water vapor which were not taken into
account in the two simple methods. To produce a more representative
reference we included the effects of these factors through variables
using a radiative transfer model (RTM). The applied methodology is
discussed in section\nobreakspace{}\ref{rtmcs}..

\hypertarget{rtmcs}{%
\subsection{Modeled clear Sky Irradiance}\label{rtmcs}}

\hypertarget{climatology-of-clear-sky-irradiance}{%
\subsubsection{Climatology of clear sky
irradiance}\label{climatology-of-clear-sky-irradiance}}

We approximated the expected clear sky \(\text{GHI}_\text{ref}\) with
the radiative transfer model uvspec, part of libRadtran
\citep{Emde2016}, similarly to the approach used by
\citet{Vamvakas2020}. In uvspec we used the solar spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), the
radiative trasfer solver ``disort'' in ``pseudospherical'' geometry and
the ``LOWTRAN'' gas parameterization. The model was run for a range of
variables in order to create a look up table (LUT) for the estimation of
the cloud-free reference irradiance for each individual observation of
our dataset. In this context, the model was run for SZAs in the range
\(10\) -- \(90^\circ\) with a step of \(0.2^\circ\) and for the
atmospheric profiles of the Air Force Geophysics Laboratory
\citep{Anderson1986} midlatitude summer and midlatitude winter,
representative of the warm and cold seasons.

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements are available since 2003 from
the Cimel sun photometer that belongs to the Aerosol Robotic Network
(AERONET) \citep{Giles2019, Buis1998}. From the observations in the
period 2003 -- 2023 we calculated the monthly climatological means and
standard deviation (\(\sigma\)) for the aerosol optical depth (AOD) at
\(500\,\text{nm}\) and the equivalent height of the water column (WC).
The monthly climatological values of AOD and water column, as well as
combinations with additional offsets of \(\pm1\sigma\) and
\(\pm2\sigma\) , were used as inputs to the RTM in the construction of
the LUT.

For each measurement of the dataset, we calculated from the LUT a
\(\text{GHI}_\text{ref}\) value for the respective season and SZA (by
linear interpolation), and for the climatological values of AOD and WC
of the respective month. The same procedure was followed for the
estimation of the \(\text{GHI}_\text{ref}\) for all combinations of the
AOD and WC with the above mentioned standard deviation offsets. Finally,
each \(\text{GHI}_\text{ref}\) value was adjusted to the actual
Sun-Earth distance derived by the Astropy software library
\citep{AstropyCollaboration2022}.

\hypertarget{long-term-change-of-clear-sky-irradiance}{%
\subsubsection{Long-term change of clear sky
irradiance}\label{long-term-change-of-clear-sky-irradiance}}

@@

\hypertarget{criteria-for-the-identification-of-ce-events}{%
\subsection{Criteria for the identification of CE
events}\label{criteria-for-the-identification-of-ce-events}}

@@

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method is capable in detecting all
major CE events. Where some CE events with very low OIR may be not
detected, these are few with small over-irradiance and it is unlikely
that will effect significantly our results.

@@

\hypertarget{results}{%
\section{Results}\label{results}}

Following the application of the above discussed methodology to the
entire dataset (6.1 million of one-minute GHI measurements),
\(1.76\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OIR of
\(345.9\,\text{W}/\text{m}^2\). The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OIR and
excess irradiation.

\hypertarget{long-term-trends}{%
\subsection{Long-term trends}\label{long-term-trends}}

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OIR and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(116.9\pm 67.8\,\text{kJ}/\text{year}\)), which are also statistically
significant. In contrast the trend of the yearly mean OIR is negligible
(\(0.1\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of no statistical
significance. The average OIR for the entire period is
\(39.9\,\text{W}/\text{m}^2 \pm 2.7\,\text{W}/\text{m}^2\). The
interannual variability of the data about the trend lines is quite
large. Furthermore, the spread tends to increase with time (at least for
the quantities of panels b and c), suggesting a significant variability
in cloud patterns over the area, possibly associated to changes in
climate.

@@

\hypertarget{climatology-of-cloud-enhancement-events}{%
\subsection{Climatology of cloud enhancement
events}\label{climatology-of-cloud-enhancement-events}}

@@

\hypertarget{duration-of-cloud-enhancement-events}{%
\subsection{Duration of cloud enhancement
events}\label{duration-of-cloud-enhancement-events}}

@@

\hypertarget{extreme-cloud-enhancement-events}{%
\subsection{Extreme cloud enhancement
events}\label{extreme-cloud-enhancement-events}}

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

By creating a clear sky approximation of the GHI, which represents the
long- and short-term variation of the expected clear sky GHI, we were
able to identify cases of CE events. After analyzing the CE cases, we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June.

The magnitude of the ECE events does not exceed the values reported for
sites with more favourable conditions for the phenomenon. The
climatological characteristic of the ECE events, showed that the most
active months are spread in half of the year (March -- June, September
and October). We found that CE conditions, can have a duration of more
than an hour in rare cases, with the bulk of the cases having a duration
under 5 minutes. Some of the characteristics of CE and ECE events we
analysed have strong similarities with results by \citet{Vamvakas2020}
for the city of Patras, south of Thessaloniki, with analog differences
on the intensity of the solar radiation.

An interpretation of the CE trends shows that the interaction of GHI
with the clouds, through this 30 year period, is a dynamic phenomenon
that needs further investigation. Although, to approach it, we need more
long-term observations of AOD and clouds characteristics.

>>>>

\section*{Abstract}\label{abstract}
\addcontentsline{toc}{section}{Abstract}

We studied the enhancement of the global horizontal irradiance induced
by clouds over Thessaloniki for the period 1993 -- 2023 using data
recorded every one minute. We identified the cloud enhancement events
(CE) by creating an appropriate clear sky irradiance reference with the
use of a radiative transfer model and aerosol optical depth data from a
collocated Cimel sun photometer and a Brewer spectrophotometer. We found
a trend in CE events of \(45.6\pm
21.9\,\text{cases}/\text{year}\), and a trend in the CE events
irradiation of \(116.9\pm
67.8\,\text{kJ}/\text{year}\). The peak of the CE events was observed
during May and June. An analysis of the total duration of CE events,
showed that durations longer than 5 minutes are very rare, with
exceptions lasting over an hour and up to 140 minutes. Finally, we have
detected enhancements above the total solar irradiance at the top of the
atmosphere of up to \(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\)
of the cases below \(200\,\text{W}/\text{m}^{2}\). The most active
season of these extreme events is spring until early summer and autumn.

\section{Introduction}\label{introduction}

@@

In this study, we evaluate the effects CE on GHI by investigating their
frequency of occurrence, intensity, and duration in a thirty-year period
of GHI observations at Thessaloniki, Greece. We used modeled clear sky
irradiance as a baseline to identify cloud enhancements, and we
determined long-term trends of the above-mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

\section{Data and methodology}\label{data-and-methodology}

\subsection{Instrumentation and data}\label{instrumentation-and-data}

@@

\subsection{Cloud enhancement
detection}\label{cloud-enhancement-detection}

@@

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}, and the total solar irradiance (TSI) at
the top of the atmosphere, adjusted for the Sun-Earth distance. We have
tested both methods by using an appropriate relative threshold and/or an
additional constant offset. The initial results showed that we can
detect a big portion of the CE events. However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric
conditions introduced numerous false positive and false negative
results. The main reason for these discrepancies is the variability of
the effects of aerosols and water vapor which were not taken into
account in the two simple methods. To produce a more representative
reference we included the effects of these factors through variables
using a radiative transfer model (RTM). The applied methodology is
discussed in section\nobreakspace{}\ref{rtmcs}.

\subsection{Modeled clear Sky Irradiance}\label{rtmcs}

\subsubsection{Climatology of clear sky
irradiance}\label{climatology-of-clear-sky-irradiance}

We approximated the expected clear sky \(\text{GHI}_\text{ref}\) with
the radiative transfer model uvspec, part of libRadtran
\citep{Emde2016}, similarly to the approach used by
\citet{Vamvakas2020}. In uvspec we used the solar spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), the
radiative transfer solver ``disort'' in ``pseudospherical'' geometry and
the ``LOWTRAN'' gas parameterization. The model was run for a range of
variables in order to create a look-up table (LUT) for the estimation of
the cloud-free reference irradiance for each individual observation of
our dataset. In this context, the model was run for SZAs in the range
\(10\) -- \(90^\circ\) with a step of \(0.2^\circ\) and for the
atmospheric profiles of the Air Force Geophysics Laboratory
\citep{Anderson1986} midlatitude summer and midlatitude winter,
representative of the warm and cold seasons.

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements are available since 2003 from
the Cimel sun photometer that belongs to the Aerosol Robotic Network
(AERONET) \citep{Giles2019, Buis1998}. From the observations in the
period 2003 -- 2023 we calculated the monthly climatological means and
standard deviation (\(\sigma\)) for the aerosol optical depth (AOD) at
\(500\,\text{nm}\) and the equivalent height of the water column (WC).
The monthly climatological values of AOD and water column, as well as
combinations with additional offsets of \(\pm1\sigma\) and
\(\pm2\sigma\), were used as inputs to the RTM in the construction of
the LUT.

For each measurement of the dataset, we calculated from the LUT a
\(\text{GHI}_\text{ref}\) value for the respective season and SZA (by
linear interpolation), and for the climatological values of AOD and WC
of the respective month. The same procedure was followed for the
estimation of the \(\text{GHI}_\text{ref}\) for all combinations of the
AOD and WC with the above-mentioned standard deviation offsets. Finally,
each \(\text{GHI}_\text{ref}\) value was adjusted to the actual
Sun-Earth distance derived by the Astropy software library
\citep{AstropyCollaboration2022}.

\subsubsection{Long-term change of clear sky
irradiance}\label{long-term-change-of-clear-sky-irradiance}

@@

\subsection{Criteria for the identification of CE
events}\label{criteria-for-the-identification-of-ce-events}

@@

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method is capable in detecting all
major CE events. Where some CE events with very low OIR may be not
detected, these are few with small over-irradiance, and it is unlikely
that will affect significantly our results.

@@

\section{Results}\label{results}

Following the application of the above discussed methodology to the
entire dataset (\(6.1\) million of one-minute GHI measurements),
\(1.76\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OIR of
\(345.9\,\text{W}/\text{m}^2\). The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OIR and
excess irradiation.

\subsection{Long-term trends}\label{long-term-trends}

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OIR and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OIR
is negligible (\(0.1\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OIR for the entire period is
\(39.9\pm 2.7\,\text{W}/\text{m}^2\). The interannual variability of the
data about the trend lines is quite large. Furthermore, the spread tends
to increase with time (at least for the quantities of panels b and c),
suggesting a significant variability in cloud patterns over the area,
possibly associated to changes in climate.

@@

\subsection{Climatology of cloud enhancement
events}\label{climatology-of-cloud-enhancement-events}

@@

\subsection{Duration of cloud enhancement
events}\label{duration-of-cloud-enhancement-events}

@@

\subsection{Extreme cloud enhancement
events}\label{extreme-cloud-enhancement-events}

@@

\section{Conclusions}\label{conclusions}

By creating a clear sky approximation of the GHI, which represents the
long- and short-term variation of the expected clear sky GHI, we were
able to identify cases of CE events. After analyzing the CE cases, we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June. We found that continuous CE events can last up to \(140\) minutes,
and \(80\,\\%\) of them bellow \(5\) minutes.

The magnitude of the ECE events does not exceed the values reported for
sites with more favourable conditions for the phenomenon. We have
observed cases exceeding the TSI by \(400\,\text{W}/\text{m}^{2}\) with
\(75\,\%\) of the cases under \(200\,\text{W}/\text{m}^{2}\). The
climatological characteristic of the ECE events, showed that the most
active months are spread in half of the year (March -- June, September
and October). We found that CE conditions, can have a duration of more
than an hour in rare cases, with the bulk of the cases having a duration
under 5 minutes. Some of the characteristics of CE and ECE events we
analysed have strong similarities with results by \citet{Vamvakas2020}
for the city of Patras, south of Thessaloniki, with analog differences
on the intensity of the solar radiation.

An interpretation of the CE trends shows that the interaction of GHI
with the clouds, through this 30-year period, is a dynamic phenomenon
that needs further investigation. Although, to approach it, we need more
long-term observations of AOD and clouds characteristics.
